---
title: Split Plot Design
author: "[Julian Faraway](https://julianfaraway.github.io/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  gfm:
    toc: true
---

```{r}
#| label: global_options
#| include: false
knitr::opts_chunk$set(comment=NA, 
                      echo = TRUE,
                      fig.path="figs/",
                      dev = 'svglite',  
                      fig.ext = ".svg",
                      warning=FALSE, 
                      message=FALSE)
knitr::opts_knit$set(global.par = TRUE)
```

```{r}
#| label: graphopts
#| include: false
par(mgp=c(1.5,0.5,0), mar=c(3.1,3.1,0.1,0), pch=20)
ggplot2::theme_set(ggplot2::theme_bw())
```

See the [introduction](index.md) for an overview. 

See a [mostly Bayesian analysis](irrigation.md) analysis of the same
data. 


This example is discussed in more detail in my book
[Extending the Linear Model with R](https://julianfaraway.github.io/faraway/ELM/)

```{r}
library(faraway)
library(ggplot2)
```

# Data

In an agricultural field trial, the
objective was to determine the effects of two crop varieties and four
different irrigation methods. Eight fields were available, but only one
type of irrigation may be applied to each field. The fields may be
divided into two parts with a different variety planted in each half.
The whole plot factor is the method of irrigation, which should be
randomly assigned to the fields. Within each field, the variety is
randomly assigned.

Load in and plot the data:

```{r}
#| label: irriplot
data(irrigation, package="faraway")
summary(irrigation)
ggplot(irrigation, aes(y=yield, x=field, shape=variety, color=irrigation)) + geom_point()
```

# Mixed Effect Model

The irrigation and variety are fixed effects, but the field is 
a random effect. We must also consider the interaction between field
and variety, which is necessarily also a random effect because one of
the two components is random. The fullest model that we might consider
is: 


$$
y_{ijk} = \mu + i_i + v_j + (iv)_{ij} + f_k + (vf)_{jk} + \epsilon_{ijk}
$$


where $\mu, i_i, v_j, (iv)_{ij}$ are fixed effects; the rest are random
having variances $\sigma^2_f$, $\sigma^2_{vf}$ and $\sigma^2_\epsilon$.
Note that we have no $(if)_{ik}$ term in this model. It would not be
possible to estimate such an effect since only one type of irrigation
is used on a given field; the factors are not crossed. Unfortunately,
it is not possible to
distinguish the variety within the field variation.  We would need
more than one observation per variety within each field for us to
separate the two variabilities. (This means that we are not
demonstrating split plot modeling in the wider sense).
We resort to a simpler model that omits the variety by field interaction random effect: 


$$
y_{ijk} = \mu + i_i + v_j + (iv)_{ij} + f_k +  \epsilon_{ijk}
$$

# LME4

See the discussion for the [single random effect example](pulpfreq.md#LME4)
for some introduction.

We fit this model with:

```{r}
library(lme4)
lmod4 = lmer(yield ~ irrigation * variety + (1|field), irrigation)
summary(lmod4, cor=FALSE)
```

The fixed effects don't look very significant. For testing
the fixed effects, we might try:

```{r}
anova(lmod4)
```

The small values of the F-statistics suggest a lack of significance
but no p-values or degrees of freedom for the error are supplied due
to previously mentioned concerns regarding correctness.

We use the `pbkrtest` package
for testing which implements the Kenward-Roger approximation for the degrees
of freedom. First test the interaction:

```{r}
library(pbkrtest)
lmoda = lmer(yield ~ irrigation + variety + (1|field),data=irrigation)
summary(lmoda, cor=FALSE)
KRmodcomp(lmod4, lmoda)
```

The interaction is not significant. Does the variety make
a difference to the yield? We fit a model with no variety effect:

```{r}
lmodi <- lmer(yield ~ irrigation + (1|field), irrigation)
```

We have some choices now. One idea is to compare this model to the
main effects only model with:

```{r}
KRmodcomp(lmoda,lmodi)
```

This assumes that the interaction effect is zero and absorbs all the
variation and degrees of freedom into the error term. Although we
did not find a significant interaction effect, this is a stronger
assumption and we may not have the best estimate of $\sigma^2$ for
the denominator in the F-test.

Another idea is to compare to the full model with:

```{r}
KRmodcomp(lmod4,lmodi)
```

The problem with this is that we are not just testing the variety effect.
This is testing the variety effect and its interaction with irrigation. That's
OK but not what we meant to test.

We can take the ANOVA approach to testing the variety but we cannot construct
this a model comparison. The `lme4` package did provide us with
the ANOVA table. We can get the (possibly adjusted) degrees of freedom
from `pbkrtest` output as 4, get the F-statistic from the ANOVA table
and compute the p-value using the F-distribution (the numerator DF also
comes from the table):

```{r}
1-pf(1.07,1,4)
```

A not significant result. In this case, all the nitpicking does not make
much difference.

Now check the irrigation method using the same ANOVA-based approach:

```{r}
1-pf(0.39,3,4)
```

We find no significant irrigation effect.


As a final check, lets compare the null model
with no fixed effects to the full model.

```{r}
lmodn <- lmer(yield ~  1 + (1|field), irrigation)
KRmodcomp(lmod4, lmodn)
```

This confirms the lack of statistical significance for the variety
and irrigation factors.


We can check the significance of the random effect (field) term with:

```{r}
library(RLRsim)
exactRLRT(lmod4)
```

We can see that there is a significant variation among the fields.

# NLME

See the discussion for the [single random effect example](pulpfreq.md#NLME)
for some introduction.

```{r}
library(nlme)
```

```{r}
nlmod = lme(yield ~ irrigation * variety, irrigation, ~ 1 | field)
summary(nlmod)
```

The estimates and standard errors are the same for the corresponding
`lme4` output. `nlme` also
p-values for the fixed effects but these are not useful to us here.
We can test the fixed with:

```{r}
anova(nlmod)
```

The result for the interaction corresponds to the calculation for `lme4` using
`pbkrtest`. The tests for the main effects are the same as the directly
calculated p-values above. This clarifies what is being tested in the ANOVA
table.

We can also take the `gls` approach with:


```{r}
gmod = gls(yield ~ irrigation * variety ,
           data= irrigation,
           correlation = corCompSymm(form = ~ 1|field))
summary(gmod)
```

We attempt a test for the fixed effects with:

```{r}
anova(gmod)
```

But the denominator DFs are wrong - should be 4 and not 8 as discussed earlier.
We could fix the p-values manually.



# MMRM

See the discussion for the [single random effect example](pulpfreq.md#MMRM)
for some introduction.

```{r}
library(mmrm)
```

As discussed in the `pulp` example, we need to create a `visit` factor
to distinguish the replicates within the fields:

```{r}
irrigation$visit = factor(rep(1:2,8))
```

```{r}
mmmod = mmrm(yield ~ irrigation * variety + cs(visit|field), irrigation)
summary(mmmod)
```

The random component is expressed as a covariance matrix. 
The SD down the diagonal will give the estimated error variance
from previous models:

```{r}
cm = mmmod$cov
sqrt(cm[1,1])
```

We can compute the correlation as:

```{r}
cm[1,2]/cm[1,1]
```

This agrees with the GLS output as expected.

We can test the fixed effects with:

```{r}
library(car)
Anova(mmmod)
```

and get the same results as previously.

# GLMMTMB

See the discussion for the [single random effect example](pulpfreq.md#GLMMTMB)
for some introduction.

```{r}
library(glmmTMB)
```

The default fit uses ML (not REML)

```{r}
gtmod <- glmmTMB(yield ~ irrigation*variety + (1|field), irrigation)
summary(gtmod)
```

This is identical with the `lme4` fit using ML.

We can use the `car` package to test the treatment effects:

```{r}
Anova(gtmod)
```

but this gives chi-square tests whereas we prefer F-tests.

# Discussion

No new issues are raised by this analysis. There are some choices
with the execution of the tests of the fixed effects but these
are not unique to this type of example. 

In the [Bayesian analyses of this data](irrigation.md), there
was more analysis of the random effects but there's not much we
can do with these in the Frequentist analyses so there's nothing
to be said.



# Package version info


```{r}
sessionInfo()
```







