---
title: Multilevel Design
author: "[Julian Faraway](https://julianfaraway.github.io/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  github_document:
    toc: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(comment=NA, 
                      echo = TRUE,
                      fig.path="figs/",
                      dev = 'svglite',  
                      fig.ext = ".svg",
                      warning=FALSE, 
                      message=FALSE)
knitr::opts_knit$set(global.par = TRUE)
```

```{r graphopts, include=FALSE}
par(mgp=c(1.5,0.5,0), mar=c(3.1,3.1,0.1,0), pch=20)
ggplot2::theme_set(ggplot2::theme_bw())
```

See the [introduction](index.md) for an overview. 

This example is discussed in more detail in my book
[Extending the Linear Model with R](https://julianfaraway.github.io/faraway/ELM/)

Required libraries:

```{r}
library(faraway)
library(ggplot2)
library(lme4)
library(pbkrtest)
library(RLRsim)
library(INLA)
library(knitr)
library(rstan, quietly=TRUE)
library(brms)
library(mgcv)
```

# Data

*Multilevel* models is a term used for models for data with hierarchical
structure. The term is most commonly used in the social sciences.
We can use the methodology we have already developed to fit some of these models.

We take as our example some data from the Junior School Project collected
from primary (U.S. term is elementary) schools in inner London. 
We math test score result from year two as the response
and try to model this as a function of gender, social class and the Raven's test
score from the first year which might be taken as a measure of ability when entering
the school. We subset the data to ignore the math scores from the first two years,
we centre the Raven score and create a combined class-by-school label:


```{r}
data(jsp, package="faraway")
jspr <- jsp[jsp$year==2,]
jspr$craven <- jspr$raven-mean(jspr$raven)
jspr$classch <- paste(jspr$school,jspr$class,sep=".")
```

We can plot the data

```{r jspplot}
ggplot(jspr, aes(x=raven, y=math))+xlab("Raven Score")+ylab("Math Score")+geom_point(position = position_jitter())
ggplot(jspr, aes(x=social, y=math))+xlab("Social Class")+ylab("Math Score")+geom_boxplot()
```

# Mixed Effect Model

As discussed in [Extending the Linear Model with R](https://julianfaraway.github.io/faraway/ELM/),
we simplify to model with an interaction between the centred Raven score and
the social class as fixed effects and the school and class nested within school as
random effects.

```{r}
mmod <- lmer(math ~ craven*social+(1|school)+(1|school:class),jspr)
faraway::sumary(mmod)
```

We can see the math score is strongly related to
the entering Raven score. We see that for the same entering score, the
final math score tends to be lower as social class goes down. 
We also see the most substantial variation at the individual level with smaller
amounts of variation at the school and class level.

We test the random effects:

```{r}
mmodc <- lmer(math ~ craven*social+(1|school:class),jspr)
mmods <- lmer(math ~ craven*social+(1|school),jspr)
exactRLRT(mmodc, mmod, mmods)
exactRLRT(mmods, mmod, mmodc)
```

The first test is for the class effect which just fails to meet the 5% significance level.
The second test is for the school effect and shows strong evidence of differences between
schools.


We can test the interaction fixed effect:

```{r}
mmodm <- lmer(math ~ craven+social+(1|school)+(1|school:class),jspr)
KRmodcomp(mmod, mmodm)
```

We see the interaction is significant.

We can compute confidence intervals for the parameters:

```{r jspconfint, cache=TRUE}
confint(mmod, method="boot")
```

The lower end of the class confidence interval is zero while the school random effect
is clearly larger.


# INLA

Integrated nested Laplace approximation is a method of Bayesian computation
which uses approximation rather than simulation. More can be found
on this topic in [Bayesian Regression Modeling with INLA](http://julianfaraway.github.io/brinla/) and the 
[chapter on GLMMs](https://julianfaraway.github.io/brinlabook/chaglmm.html)

Use the most recent computational methodology:


```{r}
inla.setOption(inla.mode="experimental")
inla.setOption("short.summary",TRUE)
```

```{r jspinladef, cache=TRUE}
formula <- math ~ social+craven + f(school, model="iid") + f(classch, model="iid")
result <- inla(formula, family="gaussian", data=jspr)
summary(result)
```

As usual, the default priors result in precisions for the random effects which are unbelievably large and we need to change the default prior.

## Informative Gamma priors on the precisions

Now try more informative gamma priors for the random effect precisions. Define it so
the mean value of gamma prior is set to the inverse of the variance of
the residuals of the fixed-effects only model. We expect the error
variances to be lower than this variance so this is an overestimate.
The variance of the gamma prior (for the precision) is controlled by
the `apar` shape parameter.

```{r jspinlaig, cache=TRUE}
apar <- 0.5
lmod <- lm(math ~ social+craven, jspr)
bpar <- apar*var(residuals(lmod))
lgprior <- list(prec = list(prior="loggamma", param = c(apar,bpar)))
formula = math ~ social+craven+f(school, model="iid", hyper = lgprior)+f(classch, model="iid", hyper = lgprior)
result <- inla(formula, family="gaussian", data=jspr)
summary(result)
```

Results are more credible.

Compute the transforms to an SD scale for the random effect terms. Make a table of summary statistics for the posteriors:

```{r sumstats}
sigmarun <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[2]])
sigmapos <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[3]])
sigmaepsilon <- inla.tmarginal(function(x) 1/sqrt(exp(x)),result$internal.marginals.hyperpar[[1]])
restab=sapply(result$marginals.fixed, function(x) inla.zmarginal(x,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmarun,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmapos,silent=TRUE))
restab=cbind(restab, inla.zmarginal(sigmaepsilon,silent=TRUE))
colnames(restab) = c("mu","B -- A","C -- A","D -- A","run","position","epsilon")
data.frame(restab) |> kable()
```

Also construct a plot the SD posteriors:

```{r plotsdsab}
ddf <- data.frame(rbind(sigmarun,sigmapos,sigmaepsilon),errterm=gl(3,nrow(sigmarun),labels = c("run","position","epsilon")))
ggplot(ddf, aes(x,y, linetype=errterm))+geom_line()+xlab("wear")+ylab("density")+xlim(0,35)
```

Posteriors look OK although no weight given to smaller values.

## Penalized Complexity Prior

In [Simpson et al (2015)](http://arxiv.org/abs/1403.4630v3), penalized complexity priors are proposed. This
requires that we specify a scaling for the SDs of the random effects. We use the SD of the residuals
of the fixed effects only model (what might be called the base model in the paper) to provide this scaling.

```{r abrainlapc, cache=TRUE}
lmod <- lm(wear ~ material, abrasion)
sdres <- sd(residuals(lmod))
pcprior <- list(prec = list(prior="pc.prec", param = c(3*sdres,0.01)))
formula = wear ~ material+f(run, model="iid", hyper = pcprior)+f(position, model="iid", hyper = pcprior)
result <- inla(formula, family="gaussian", data=abrasion)
summary(result)
```

Compute the summaries as before:

```{r ref.label="sumstats"}
```

Make the plots:

```{r abrapc, ref.label="plotsdsab"}
```

Posteriors put more weight on lower values compared to gamma prior.

# STAN

[STAN](https://mc-stan.org/) performs Bayesian inference using
MCMC.
Set up STAN to use multiple cores. Set the random number seed for reproducibility.

```{r}
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(123)
```

Fit the model. Requires use of STAN command file [abrasion.stan](../stancode/abrasion.stan). We have used uninformative priors for the fixed effects and half-cauchy priors for the three variances. Prepare data in a format consistent with the command file. Needs to be a list:

```{r}
sdscal <- sd(residuals(lm(wear ~ material, abrasion)))
abrdat <- list(N=16, Nt=4, treat=as.numeric(abrasion$material), blk1=as.numeric(abrasion$run), blk2=as.numeric(abrasion$position), y=abrasion$wear, sdscal=sdscal)
```

Run in three steps:

```{r abrastan, cache=TRUE}
rt <- stanc(file="../stancode/abrasion.stan")
sm <- stan_model(stanc_ret = rt, verbose=FALSE)
system.time(fit <- sampling(sm, data=abrdat))
```

We have not used an overall mean. If we want an overall mean
parameter, we have to set up dummy variables. We can do this but it
requires more work.


## Diagnostics

For the error SD:

```{r abrasigmaepsdiag}
pname <- "sigmaeps"
muc <- rstan::extract(fit, pars=pname,  permuted=FALSE, inc_warmup=FALSE)
mdf <- reshape2::melt(muc)
ggplot(mdf,aes(x=iterations,y=value,color=chains)) + geom_line() + ylab(mdf$parameters[1])
```

For the block one (run) SD:

```{r abrasigmab1diag}
pname <- "sigmab1"
muc <- rstan::extract(fit, pars=pname,  permuted=FALSE, inc_warmup=FALSE)
mdf <- reshape2::melt(muc)
ggplot(mdf,aes(x=iterations,y=value,color=chains)) + geom_line() + ylab(mdf$parameters[1])
```

For the block two (position) SD:

```{r abrasigmab2}
pname <- "sigmab2"
muc <- rstan::extract(fit, pars=pname,  permuted=FALSE, inc_warmup=FALSE)
mdf <- reshape2::melt(muc)
ggplot(mdf,aes(x=iterations,y=value,color=chains)) + geom_line() + ylab(mdf$parameters[1])
```

Everything looks reasonable.

## Output Summary

```{r}
print(fit,pars=c("trt","sigmaeps","sigmab1","sigmab2"))
```

The effective sample sizes are a bit low so we might want to rerun this with more iterations
if we care about the tails in particular.

## Posterior Distributions

We can use extract to get at various components of the STAN fit. First consider the SDs for random components:

```{r abrapostsig}
postsig <- rstan::extract(fit, pars=c("sigmaeps","sigmab1","sigmab2"))
ref <- reshape2::melt(postsig)
colnames(ref)[2:3] <- c("wear","SD")
ggplot(data=ref,aes(x=wear, color=SD))+geom_density()
```

As usual the error SD distribution is a bit more concentrated. We can see
some weight at zero for the random effects in contrast to the INLA posteriors.

Now the treatment effects:


```{r abraposttrt}
postsig <- rstan::extract(fit, pars="trt")
ref <- reshape2::melt(postsig)
colnames(ref)[2:3] <- c("material","wear")
ref$material <- LETTERS[1:4][ref$material]
ggplot(data=ref,aes(x=wear, color=material))+geom_density()
```

We can see that material A shows some separation from the other levels.

# BRMS

[BRMS](https://paul-buerkner.github.io/brms/) stands for Bayesian Regression Models with STAN. It provides a convenient wrapper to STAN functionality. We specify the model as in `lmer()` above.
I have used more than the standard number of iterations because this reduces some problems
and does not cost much computationally.


```{r brmfit, cache=TRUE}
suppressMessages(bmod <- brm(wear ~ material + (1|run) + (1|position), data=abrasion,iter=10000, cores=4))
```

We get some minor warnings. We can obtain some posterior densities and diagnostics with:

```{r eggsbrmsdiag}
plot(bmod, variable = "^s", regex=TRUE)
```

We have chosen only the random effect hyperparameters since this is
where problems will appear first. Looks OK. We can see some weight
is given to values of the random effect SDs close to zero.

We can look at the STAN code that `brms` used with:

```{r}
stancode(bmod)
```

We see that `brms` is using student t distributions with 3 degrees of
freedom for the priors. For the three error SDs, this will be truncated at
zero to form half-t distributions. You can get a more explicit description
of the priors with `prior_summary(bmod)`. These are qualitatively similar to the
the PC prior used in the INLA fit. 

We examine the fit:

```{r}
summary(bmod)
```

The results are consistent with those seen previously.

# MGCV

It is possible to fit some GLMMs within the GAM framework of the `mgcv`
package. An explanation of this can be found in this 
[blog](https://fromthebottomoftheheap.net/2021/02/02/random-effects-in-gams/)


```{r}
gmod = gam(wear ~ material + s(run,bs="re") + s(position,bs="re"),
           data=abrasion, method="REML")
```

and look at the summary output:

```{r}
summary(gmod)
```

We get the fixed effect estimates.
We also get tests on the random effects (as described in this [article](https://doi.org/10.1093/biomet/ast038). The hypothesis of no variation
is rejected for both the run and the position. This is consistent
with earlier findings.

We can get an estimate of the operator and error SD:

```{r}
gam.vcomp(gmod)
```

The point estimates are the same as the REML estimates from `lmer` earlier.
The confidence intervals are different. A bootstrap method was used for
the `lmer` fit whereas `gam` is using an asymptotic approximation resulting
in substantially different results. Given the problems of parameters on
the boundary present in this example, the bootstrap results appear more
trustworthy.

The random effect estimates for the fields can be found with:

```{r}
coef(gmod)
```

# GINLA

In [Wood (2019)](https://doi.org/10.1093/biomet/asz044), a
simplified version of INLA is proposed. The first
construct the GAM model without fitting and then use
the `ginla()` function to perform the computation.

```{r}
gmod = gam(wear ~ material + s(run,bs="re") + s(position,bs="re"),
           data=abrasion, fit = FALSE)
gimod = ginla(gmod)
```

We get the posterior density for the intercept as:

```{r abraginlaint}
plot(gimod$beta[1,],gimod$density[1,],type="l",xlab="yield",ylab="wear")
```

and for the material effects as:

```{r abraginlalmat}
xmat = t(gimod$beta[2:4,])
ymat = t(gimod$density[2:4,])
matplot(xmat, ymat,type="l",xlab="wear",ylab="density")
legend("right",paste0("Material",LETTERS[2:4]),col=1:3,lty=1:3)
```

We can see some overlap between the effects but clear separation from zero.

The run effects are:

```{r abraginlalrun}
xmat = t(gimod$beta[5:8,])
ymat = t(gimod$density[5:8,])
matplot(xmat, ymat,type="l",xlab="wear",ylab="density")
legend("right",paste0("run",1:4),col=1:4,lty=1:4)
```

All the run effects overlap with zero but runs 3 and 4 are more distinct than 1 and 2.

The position effects are:

```{r abraginlalpos}
xmat = t(gimod$beta[9:12,])
ymat = t(gimod$density[9:12,])
matplot(xmat, ymat,type="l",xlab="wear",ylab="density")
legend("right",paste0("position",1:4),col=1:4,lty=1:4)
```

Here positions 1 and 2 are more distinct in their effects.

It is not straightforward to obtain the posterior densities of
the hyperparameters. 

# Discussion

See the [Discussion of the single random effect model](pulp.md#Discussion) for
general comments. 

- All but the INLA analysis have some ambiguity about whether there is much difference in the
run and position effects. In the mixed effect model, the effects are just statistically significant
while the Bayesian analyses also suggest these effects have some impact while still expressing
some chance that they do not. The INLA analysis does not give any weight to the no effect claim.

- None of the analysis had any problem with the crossed effects.

- There were no major computational issue with the analyses (in contrast with some of
the other examples)

# Package version info

```{r}
sessionInfo()
```



