---
title: Poisson GLMM
author: "[Julian Faraway](https://julianfaraway.github.io/)"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  gfm:
    toc: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(comment=NA, 
                      echo = TRUE,
                      fig.path="figs/",
                      dev = 'svglite',  
                      fig.ext = ".svg",
                      warning=FALSE, 
                      message=FALSE)
knitr::opts_knit$set(global.par = TRUE)
```

```{r graphopts, include=FALSE}
par(mgp=c(1.5,0.5,0), mar=c(3.1,3.1,0.1,0), pch=20)
ggplot2::theme_set(ggplot2::theme_bw())
```


See the [introduction](../index.md) for an overview. 

This example is discussed in more detail in the book
[Bayesian Regression Modeling with INLA](https://julianfaraway.github.io/brinlabook/chaglmm.html#sec:poissonglmm)

Packages used:

```{r}
library(ggplot2)
library(lme4)
library(INLA)
library(knitr)
library(brms)
library(mgcv)
```


# Data and Model

In [Davison and Hinkley, 1997]( https://doi.org/10.1017/CBO9780511802843), the results of a study on Nitrofen, a herbicide, are reported. Due to concern regarding the effect on animal life, 50 female water fleas were divided into five groups of ten each and treated with different concentrations of the herbicide. The number of offspring in three subsequent broods for each flea was recorded. We start by loading the data from the `boot` package: (the `boot` package comes with base R distribution  so there is no need to download this)

```{r}
#| label: nitrofendat
data(nitrofen, package="boot")
head(nitrofen)
```

We need to rearrange the data to have one response value per line:

```{r}
lnitrofen = data.frame(conc = rep(nitrofen$conc,each=3),
  live = as.numeric(t(as.matrix(nitrofen[,2:4]))),
  id = rep(1:50,each=3),
  brood = rep(1:3,50))
head(lnitrofen)
```

Make a plot of the data:

```{r}
#| label: fig-nitrodat
#| fig-cap: "The number of live offspring varies with the concentration of Nitrofen and the brood number."
lnitrofen$jconc <- lnitrofen$conc + rep(c(-10,0,10),50)
ggplot(lnitrofen, aes(x=jconc,y=live, shape=factor(brood))) + 
       geom_point(position = position_jitter(w = 0, h = 0.5)) + 
       xlab("Concentration") + labs(shape = "Brood")
```

Since the response is a small count, a Poisson model is a natural
choice. We expect the rate of the response to vary with the brood and
concentration level. The plot of the data suggests these two predictors
may have an interaction. The three observations for a single flea are
likely to be correlated. We might expect a given flea to tend to produce
more, or less, offspring over a lifetime. We can model this with an
additive random effect. The linear predictor is:
$$
\eta_i = x_i^T \beta + u_{j(i)}, \quad i=1, \dots, 150. \quad j=1, \dots 50,
$$
where $x_i$ is a vector from the design matrix encoding the information
about the $i^{th}$ observation and $u_j$ is the random affect associated
with the $j^{th}$ flea. The response has distribution
$Y_i \sim Poisson(\exp(\eta_i))$.

# LME4

We fit a model using penalized quasi-likelihood
(PQL) using the `lme4` package:
```{r}
glmod <- glmer(live ~ I(conc/300)*brood + (1|id), nAGQ=25, 
             family=poisson, data=lnitrofen)
summary(glmod, correlation = FALSE)
```
We scaled the concentration by dividing by 300 (the maximum value is
310) to avoid scaling problems encountered with `glmer()`. This is
helpful in any case since it puts all the parameter estimates on a
similar scale. The first brood is the reference level so the slope for
this group is estimated as $-0.0437$ and is not statistically
significant, confirming the impression from the plot. We can see that
numbers of offspring in the second and third broods start out
significantly higher for zero concentration of the herbicide, with
estimates of $1.1688$ and $1.3512$. But as concentration increases, we
see that the numbers decrease significantly, with slopes of $-1.6730$
and $-1.8312$ relative to the first brood. The individual SD is
estimated at $0.302$ which is noticeably smaller than the estimates
above, indicating that the brood and concentration effects outweigh the
individual variation.

We can make a plot of the mean predicted response as concentration and
brood vary. I have chosen not to include any random effects with
the option `re.form=~0` because we are not interested in a particular `id`.


```{r}
#| label: fig-prednitro
#| fig-cap: "Predicted number of live offspring"
predf = data.frame(conc=rep(c(0,80,160,235,310),each=3),brood=rep(1:3,5))
predf$live = predict(glmod, newdata=predf, re.form=~0, type="response")
predf$brood = factor(predf$brood)
ggplot(predf, aes(x=conc,y=live,group=brood,color=brood)) + 
  geom_line() + xlab("Concentration")
```

We see that if only the first brood were considered, the herbicide does not
have a large effect. In the second and third broods, the (negative) effect of the herbicide becomes more apparent with fewer live offspring being produced
as the concentration rises.

# INLA

Integrated nested Laplace approximation is a method of Bayesian computation
which uses approximation rather than simulation. More can be found
on this topic in [Bayesian Regression Modeling with INLA](http://julianfaraway.github.io/brinla/) and the 
[chapter on GLMMs](https://julianfaraway.github.io/brinlabook/chaglmm.html)

The same model, with default priors, can be fitted with INLA as:
```{r}
#| label: nitroinla1
#| cache: true
formula <- live ~ I(conc/300)*brood + f(id, model="iid")
imod <- inla(formula, family="poisson", data=lnitrofen)
```
The fixed effects summary is:
```{r}
imod$summary.fixed |> kable()
```
The posterior means are very similar to the PQL estimates. We
can get plots of the posteriors of the fixed effects:
```{r}
#| label: fig-nitrofpd
#| fig-cap: "Posterior densities of the fixed effects model for the Nitrofen data."
fnames = names(imod$marginals.fixed)
par(mfrow=c(2,2))
for(i in 1:4){
  plot(imod$marginals.fixed[[i]],
       type="l",
       ylab="density",
       xlab=fnames[i])
  abline(v=0)
}
par(mfrow=c(1,1))
```

We can also
see the summary for the random effect SD:
```{r}
hpd = inla.tmarginal(function(x) 1/sqrt(x), imod$marginals.hyperpar[[1]])
inla.zmarginal(hpd)
```
Again the result is very similar to the PQL output although notice that
INLA provides some assessment of uncertainty in this value in contrast
to the PQL result. We can also see the posterior density:

```{r}
#| label: fig-nitrohyppd
#| fig-cap: "Posterior density of the SD of id"
plot(hpd,type="l",xlab="linear predictor",ylab="density")
```


# BRMS

For this example, I did not write my own STAN program. I am not
that experienced in writing STAN programmes so it is better rely
on the superior experience of others.

[BRMS](https://paul-buerkner.github.io/brms/) stands for Bayesian Regression Models with STAN. It provides a convenient wrapper to STAN functionality.

Fitting the model is very similar to `lmer` as seen above:

```{r}
#| label: brmfit
#| cache: true
suppressMessages(bmod <- brm(live ~ I(conc/300)*brood + (1|id),  
             family=poisson, data=lnitrofen))
```

We can check the MCMC diagnostics and the posterior densities with:

```{r}
#| label: fig-nitrobrmsdiag
plot(bmod)
```

Looks quite similar to the INLA results.

We can look at the STAN code that `brms` used with:

```{r}
stancode(bmod)
```

We can see that some half-t distributions are used as priors for
the hyperparameters.

We examine the fit:

```{r}
summary(bmod)
```

The results are consistent with previous results.

# MGCV

It is possible to fit some GLMMs within the GAM framework of the `mgcv`
package. An explanation of this can be found in this 
[blog](https://fromthebottomoftheheap.net/2021/02/02/random-effects-in-gams/)

We need to make a factor version of id otherwise it gets treated
as a numerical variable.

```{r}
lnitrofen$fid = factor(lnitrofen$id)
gmod = gam(live ~ I(conc/300)*brood + s(fid,bs="re"), 
           data=lnitrofen, family="poisson", method="REML")
```

and look at the summary output:

```{r}
summary(gmod)
```

We get the fixed effect estimates.
We also get a test on the random effect (as described in this [article](https://doi.org/10.1093/biomet/ast038)). The hypothesis of no variation
between the ids is rejected.

We can get an estimate of the id SD:

```{r}
gam.vcomp(gmod)
```

which is the same as the REML estimate from `lmer` earlier.

The random effect estimates for the fields can be found with:

```{r}
coef(gmod)
```

# GINLA

In [Wood (2019)](https://doi.org/10.1093/biomet/asz044), a
simplified version of INLA is proposed. The first
construct the GAM model without fitting and then use
the `ginla()` function to perform the computation.

```{r}
gmod = gam(live ~ I(conc/300)*brood + s(fid,bs="re"), 
           data=lnitrofen, family="poisson", fit = FALSE)
gimod = ginla(gmod)
```

We get the posterior densities for the fixed effects as:

```{r}
#| label: fig-nitroginlareff
#| fig-cap: "Posteriors of the fixed effects"
par(mfrow=c(2,2))
for(i in 1:4){
plot(gimod$beta[i,],gimod$density[i,],type="l",
     xlab=gmod$term.names[i],ylab="density")
}
par(mfrow=c(1,1))
```

It is not straightforward to obtain the posterior densities of
the hyperparameters. 

# Discussion

- No strong differences in the results between the different methods.

- LME4, MGCV and GINLA were very fast. INLA was fast. BRMS was slowest. But
this is a small dataset and a simple model so we cannot draw too general
a conclusion from this.

# Package version info


```{r}
sessionInfo()
```

